### 2018-01-05 submission/manuscript.tex
results re-write upto diversity before CRC-associated re-analysis

\textbf{\emph{Carcinoma-Associated Genera Minimally Impacts RR of
Adenoma:}} Based on the small increase in RR using \(\alpha\)-diversity
metrics, we assessed if the presence of specific genera resulted in a
higher RR for both stool and tissue. To investigate this we analyzed the
classically associated CRC genera, \emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, and \emph{Porphyromonas}
for an increase in RR. The majority of CRC-associated genera for both
feces and tissue had a significantly increased RR for carcinoma but not
for adenoma {[}Figure 3{]}. The RR effect size was greater for stool (RR
range = 1.62 - 2.37) than for tissue (RR range = 1.21 - 1.81). This
decrease may be explained by the fact that the tissue analysis included
matched samples from the same individual. In fecal samples, the RR for
carcinoma due to the presence of CRC-associated genera was greater than
either the RR associated with evenness or Shannon diversity {[}Figure 2
\& 3{]}. Additionally, the RR of carcinoma continuously increased as
individuals tested positive for more CRC-associated genera {[}Figure 3B
\& 3D{]}.

There were two significant measures for increased RR of adenoma when
investigating CRC-associated genera in stool: 1) Having a higher then
median value of \emph{Porphyromonas} (P-value = 0.023) and 2) whether
samples were positive for three CRC-associated genera (P-value = 0.022)
{[}Figure 3A{]}. With tissue, there were three significant measures for
an increased RR of adenoma: 1) being positive for one CRC-associated
genera (P-value = 0.032), 2) being positive for two CRC-associated
genera (P-value = 0.008), and 3) being positive for four CRC associated
genera (P-value = 0.039) {[}Figure 3C{]}.

\textbf{\emph{Whole Community Models Add Important Community Context:}}
Since CRC-associated genera had increased RR for carcinoma over
diversity metrics we wondered whether the overall bacterial community
was at all important to classifying disease or if the CRC-associated
genera were sufficient alone. To test this we used two approaches. The
first approach used genus level data and tested whether there were any
differences in AUC when training on one study and testing on all the
others when using either all genera present or only the CRC-associated
genera. The second approach used OTU level data and tested whether there
was a generalized decrease in the 10-fold cross validation (CV) model
across studies using either all OTUs or only OTUs that taxonomically
classified to CRC-associated genera.

Our first approach using genus based models showed an AUC decrease in
model classification on the training set for both stool and tissue
studies {[}Figure S2-S3{]}. With respect to the test sets, comprised of
genera data from other studies, both the all genera model and
CRC-associated models had a similar ability to detect adenomas or
carcinomas {[}Figure S4-S6{]}. Two interesting but separate general
observations from these models were that: 1) classification of adenomas
was lower than carcinomas for both tissue and stool and 2) AUC for the
classification of carcinoma was consistently lower for the tissue models
than the stool models {[}Figure S4-S6{]}.

A similar trend was observed for the OTU based models for both fecal and
tissue (matched and unmatched) samples. There was a generalized decrease
in AUC when only OTUs from the CRC-associated genera were used versus
the full community of OTUs for both adenoma and carcinoma {[}Figure 4 \&
5{]}. The largest difference in median AUC for stool, between the full
community of OTUs and only CRC-associated OTUs, was in carcinoma
classification {[}Figure 4B{]} and for tissue it was in adenoma
classification {[}Figure 5A{]}.

In stool the most common genera in the genus based models belonged
mostly to resident genera such as \emph{Ruminococcus},
\emph{Bacteroides}, and \emph{Roseburia} {[}Figure 6A \& B{]}. With
respect to the CRC-associated genera, \emph{Fusobacterium} was the only
genus present in adenoma while all four were present in carcinoma
{[}Figure 6A \& B{]}. Conversely, none of these CRC-associated genera
were present in the majority of studies. When we move on to the OTU
based models, the adenoma OTU models had OTUs that classified as
\emph{Ruminococcaceae} or \emph{Roseburia} present in the top 10 OTUs
for the vast majority of studies {[}Figure 6C{]}. \emph{Ruminococcaceae}
was also present in the top 10 in some studies for the carcinoma OTU
models, but it was \emph{Bacteroides} that was present in the
overwhelming majority of the carcinoma OTU stool models {[}Figure 6D{]}.

Unlike stool, tissue based Random Forest models showed no consistent
representation of \emph{Ruminococcaceae}, \emph{Ruminococcus},
\emph{Bacteroides}, and \emph{Roseburia} in the top 10 across study
{[}Figure S7{]}. The vast majority of the top 10 genera and OTUs occured
in a study specific manner for these tissue based Random Forest models.
For both the genera and OTU Random Forest models of adenoma and
carcinoma, there appears to be very little overlap in the top 10 most
important variables between stool and tissue {[}Figure 6 \& S7{]}. This
discordance between stool and tissue also applies to the
mouth-associated genera with one noticeable skew being that
\emph{Fusobacterium} and \emph{Fusobacteriaceae} occur more often in the
top 10 of Random Forest models using matched versus unmatched tissue
samples {[}Figure S7B-C \& S7E-F{]}.

@@

Our study identifies clear differences in diversity, both at the
community level and for individual genera, that are present in patients
with and without CRC {[}Figure 1-3{]}. Although there was a step-wise
decrease in diversity from control to adenoma to carcinoma, this did not
translate into large effect sizes for the RR of lesion. Even though
mouth-associated genera increased the RR of carcinoma, they did not
consistently increase the RR of adenoma. These mouth-associated genera
are clearly important to carcinoma classification but ourobservations
suggest that accounting for the community in which these microbes exist
can increase the ability of models to make predictions.

The data presented herein supports the importance of mouth-associated
genera for carcinoma but not necessarily adenoma formation. A proposed
mechanism by which these mouth-associated microbes cause CRC stems from
the creation of a vicious cycle involving host inflammation and repair.
Both the drastically increased carcinoma RR of mouth-associated genera
versus \(\alpha\)-diversity metrics and increasing RR with more
mouth-associated genera positivity {[}Figure 3C{]}, are highly
supportive of their importance. However, our observations show that when
adding the community context to our Random Forest models along with the
mouth-associated genera, the model AUC increases {[}Figure 4 \& 5{]}.
This suggests the resident community cannot be ignored completely and
needs to be accounted for when investigating mouth-associated microbes
and carcinoma. Conversely, using the present data, it is less likely
that mouth-associated microbes are correlated with adenoma development.
The step-wise decrease in diversity suggests that the adenoma community
is not normal but has changed subtly {[}Figure 1{]}. Yet, this change in
diversity, at this early stage of disease, could be focal to the adenoma
itself. The poor performance of the Random Forest models for classifying
adenoma, based only on the stool microbiota, would suggest that this is
the case. One potential hypothesis from these observations is that at
early stages of the disease, how the host interacts with these subtle
changes is what ultimately leads to a thoroughly dysfunctional community
that is supportive of CRC genesis.

Within stool, common resident microbes were most consistently present in
the top 10 genera or OTUs across study {[}Figure 6{]}. Changes in
\emph{Bacteroides}, \emph{Ruminococcaceae}, \emph{Ruminococcus}, and
\emph{Roseburia} were consistently found to be disciminative across the
different studies for both adenoma and carcinoma {[}Figure 6{]}. This
data would suggest that whether the non-resident bacterium is
\emph{Fusobacteria} or \emph{Peptostreptococcus} is not as important as
how these bacteria interact with the changing resident community. Based
on these observations, it is possible to hypothesize that tsmall changes
in community structure lead to new niches in which any one of the
mouth-associated genera can gain a foothold, exacerbating the initial
changes in community and facilitating the transition from adenoma to
carcinoma.

The tissue studies did not provide a clearer understanding of how the
microbiota may be associated with lesion. Generally, the OTU models of
unmatched and matched {[}Figure S7E \& F{]} tissue samples were
concordant with stool showing that resident microbes were the most
prevalent in the top 10 across study. Unlike in stool,
\emph{Fusobacterium} was the only mouth-associated bacteria consistently
present in the top 10 of the CRC models {[}Figure S7B-C \& E-F{]}. The
majority of the tissue results seem to be study specific with many top
10 taxa being present only in a single study. Additionally, the presence
of \emph{Propionibacterium}, within the top 10 most important variables
for the genera and OTU models, is worrying because it is commonly a
marker of contamination. The low bacterial biomass of tissue samples
coupled with potential contaminationcould ultimately explain why the
these results seem to be more sporadic than the stool results.

@@

\textbf{\emph{Study Analysis Overview:}} \(\alpha\)-diversity was first
assessed for differences between controls, adenoma, and carcinoma using
both linear mixed-effect models and RR. \(\beta\)-diversity was then
assessed for each individual study for differences between
control-adenoma and control-carcinoma. Next, four specific
mouth-associated genera (\emph{Fusobacterium}, \emph{Parvimonas},
\emph{Peptostreptococcus}, and \emph{Porphyromonas}) were assessed for
differences in RR for adenoma and carcinoma. We then built Random Forest
models based on all or mouth-associated genera and these models were
trained on one study then tested on the remaining studies, for every
study. A similar approach was then applied at the OTU level with the
exception that a 10-fold CV over 100 different models, based on random
80/20 splitting of the data, was used to generate a range of expected
AUCs. For these OTU based models, the mouth-associated genera included
all OTUs that had a taxonomic classification to \emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas}.
Finally, the power of each study was assessed for an effect size ranging
from 1\% to 30\% and an estimated sample size, for these effect sizes,
was generated based on 80\% power. For comparisons in which normal
versus adenoma were made the carcinoma samples were excluded from each
respective study. Similarly, for comparisons in which normal versus
carcinoma were made the adenoma samples were excluded from each
respective study. The data was split between feces and tissue samples.
Within the tissue groups the data was further divided between samples
from the same individual (matched) and those from different individuals
(unmatched). Where applicable for each study, predictions for adenoma
and carcinoma were then tested for feces, matched tissue, and unmatched
tissue.

\textbf{\emph{Obtaining Mouth-Associated Genera:}} For the
mouth-associated genera analysis of the RR, the total average counts,
for 100 different subsamplings, were collected for each respective OTU
that had a genus level taxonomic classification to \emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, and \emph{Porphyromonas}.
The OTU based Random Forest Models using mouth-associated genera
utilized a similar approach except that the OTUs were not aggregated
together by genus but kept as separate OTUs. OTU Random Forest models
using the full community included all OTUs while those using
mouth-associated genera included only those OTUs that had a genus level
taxonomic classification to \emph{Fusobacterium}, \emph{Parvimonas},
\emph{Peptostreptococcus}, and \emph{Porphyromonas}.

>>>>

\textbf{\emph{Carcinoma-Associated Genera Minimally Impacts RR of
Adenoma:}} Based on the small increase in RR using \(\alpha\)-diversity
metrics, we assessed if the presence of specific genera resulted in a
higher RR for both stool and tissue. To investigate this we analyzed the
classically associated CRC genera, \emph{Bacteroides},
\emph{Escherichia}, \emph{Fusobacterium}, \emph{Parvimonas},
\emph{Peptostreptococcus}, \emph{Porphyromonas}, and
\emph{Streptococcus} for an increase in RR. The majority of
CRC-associated genera for both feces and tissue had a significantly
increased RR for carcinoma but not for adenoma {[}Figure 3{]}. The RR
effect size was greater for stool (RR range = 1.62 - 2.37) than for
tissue (RR range = 1.21 - 1.81). This decrease may be explained by the
fact that the tissue analysis included matched samples from the same
individual. In fecal samples, the RR for carcinoma when having
CRC-associated genera at a higher than overal study median relative
abundance was greater than the RR associated with evenness or Shannon
diversity {[}Figure 2 \& 3{]}. Additionally, the RR of carcinoma
continuously increased as individuals tested positive for more
CRC-associated genera {[}Figure 3B \& 3D{]}.

In stool the significant measures that increased RR of adenoma for
CRC-associated genera were having a higher than overall study median
relative abundance of \emph{Porphyromonas} (P-value = 0.023) and being
positive for three CRC-associated genera (P-value = 0.022) {[}Figure
3A{]}. Within tissue the significant measures for an increased RR of
adenoma were being positive for one CRC-associated genera (P-value =
0.032), being positive for two CRC-associated genera (P-value = 0.008),
and being positive for four CRC associated genera (P-value = 0.039)
{[}Figure 3C{]}.

\textbf{\emph{Whole Community Models Add Important Context:}} Since
CRC-associated genera increased RR for carcinoma over diversity metrics
we assessed whether the bacterial community was better at classifying
disease versus only CRC-associated genera and tested this using two
approaches. The first approach used genus level data and tested whether
there were any differences in AUC between all genera and CRC-associated
genera only during training on one study or testing on all other
studies. The second approach used OTU level data and tested whether
there was a generalized decrease in the 10-fold cross validation (CV)
model AUC across studies when comparing all OTUs versus only OTUs that
taxonomically classified to CRC-associated genera.

For the genera-based models both tissue and stool studies had a median
AUC decrease in model classification when only CRC-associated genera
versus all genera were included {[}Figure S2-S3{]}. With respect to the
test sets, comprised of genera data from other studies, both all and
CRC-associated genera models had a similar ability to detect adenomas or
carcinomas {[}Figure S4-S6{]}. The OTU-based models showed a similar
trend to the genera models where a generalized decrease in median AUC
was observed when only OTUs from the CRC-associated genera were used
versus the full community of OTUs for both adenoma and carcinoma, with
the largest median change for stool in carcinoma models and for tissue
in adenoma models {[}Figure 4 \& 5{]}.

In stool the most common genera in the genus based models belonged to
genera such as \emph{Ruminococcus}, \emph{Bacteroides}, and
\emph{Roseburia} {[}Figure 6A \& B{]}. Although all four CRC-associated
genera were present in carcinoma, none were present in the majority of
studies and \emph{Fusobacterium} was the only genus present in adenoma
{[}Figure 6A \& B{]}. For the OTU based models, \emph{Ruminococcaceae}
was present in the top 10 consistently for both adenoma and carcinoma
models while \emph{Roseburia} was only present in many adenoma models
and \emph{Bacteroides} was present in the overwhelming majority of the
carcinoma models {[}Figure 6C \& 6D{]}.

Unlike stool, tissue based Random Forest models showed no consistent
representation of \emph{Ruminococcaceae}, \emph{Ruminococcus},
\emph{Bacteroides}, and \emph{Roseburia} in the top 10 across study
{[}Figure S7{]}. The vast majority of the top 10 genera and OTUs occured
in a study specific manner for these tissue based Random Forest models.
For both the genera and OTU Random Forest models of adenoma and
carcinoma, there appears to be very little overlap in the top 10 most
important variables between stool and tissue {[}Figure 6 \& S7{]}. This
discordance between stool and tissue also applies to the CRC-associated
genera with one noticeable skew being that \emph{Fusobacterium} and
\emph{Fusobacteriaceae} occur more often in the top 10 of Random Forest
models using matched versus unmatched tissue samples {[}Figure S7B-C \&
S7E-F{]}.

@@

Our study identifies clear differences in diversity, both at the
community level and for individual genera, that are present in patients
with and without CRC {[}Figure 1-3{]}. Although there was a step-wise
decrease in diversity from control to adenoma to carcinoma, this did not
translate into large effect sizes for the RR of lesion. Even though
CRC-associated genera increased the RR of carcinoma, they did not
consistently increase the RR of adenoma. These CRC-associated genera are
clearly important to carcinoma classification but ourobservations
suggest that accounting for the community in which these microbes exist
can increase the ability of models to make predictions.

The data presented herein supports the importance of CRC-associated
genera for carcinoma but not necessarily adenoma formation. A proposed
mechanism by which these CRC-associated microbes cause CRC stems from
the creation of a vicious cycle involving host inflammation and repair.
Both the drastically increased carcinoma RR of CRC-associated genera
versus \(\alpha\)-diversity metrics and increasing RR with more
CRC-associated genera positivity {[}Figure 3C{]}, are highly supportive
of their importance. However, our observations show that when adding the
community context to our Random Forest models along with the
CRC-associated genera, the model AUC increases {[}Figure 4 \& 5{]}. This
suggests the resident community cannot be ignored completely and needs
to be accounted for when investigating CRC-associated microbes and
carcinoma. Conversely, using the present data, it is less likely that
CRC-associated microbes are correlated with adenoma development. The
step-wise decrease in diversity suggests that the adenoma community is
not normal but has changed subtly {[}Figure 1{]}. Yet, this change in
diversity, at this early stage of disease, could be focal to the adenoma
itself. The poor performance of the Random Forest models for classifying
adenoma, based only on the stool microbiota, would suggest that this is
the case. One potential hypothesis from these observations is that at
early stages of the disease, how the host interacts with these subtle
changes is what ultimately leads to a thoroughly dysfunctional community
that is supportive of CRC genesis.

Within stool, common resident microbes were most consistently present in
the top 10 genera or OTUs across study {[}Figure 6{]}. Changes in
\emph{Bacteroides}, \emph{Ruminococcaceae}, \emph{Ruminococcus}, and
\emph{Roseburia} were consistently found to be disciminative across the
different studies for both adenoma and carcinoma {[}Figure 6{]}. This
data would suggest that whether the non-resident bacterium is
\emph{Fusobacteria} or \emph{Peptostreptococcus} is not as important as
how these bacteria interact with the changing resident community. Based
on these observations, it is possible to hypothesize that tsmall changes
in community structure lead to new niches in which any one of the
CRC-associated genera can gain a foothold, exacerbating the initial
changes in community and facilitating the transition from adenoma to
carcinoma.

The tissue studies did not provide a clearer understanding of how the
microbiota may be associated with lesion. Generally, the OTU models of
unmatched and matched {[}Figure S7E \& F{]} tissue samples were
concordant with stool showing that resident microbes were the most
prevalent in the top 10 across study. Unlike in stool,
\emph{Fusobacterium} was the only CRC-associated bacteria consistently
present in the top 10 of the CRC models {[}Figure S7B-C \& E-F{]}. The
majority of the tissue results seem to be study specific with many top
10 taxa being present only in a single study. Additionally, the presence
of \emph{Propionibacterium}, within the top 10 most important variables
for the genera and OTU models, is worrying because it is commonly a
marker of contamination. The low bacterial biomass of tissue samples
coupled with potential contaminationcould ultimately explain why the
these results seem to be more sporadic than the stool results.

@@

\textbf{\emph{Study Analysis Overview:}} \(\alpha\)-diversity was first
assessed for differences between controls, adenoma, and carcinoma using
both linear mixed-effect models and RR. \(\beta\)-diversity was then
assessed for each individual study for differences between
control-adenoma and control-carcinoma. Next, four specific
CRC-associated genera (\emph{Fusobacterium}, \emph{Parvimonas},
\emph{Peptostreptococcus}, and \emph{Porphyromonas}) were assessed for
differences in RR for adenoma and carcinoma. We then built Random Forest
models based on all or CRC-associated genera and these models were
trained on one study then tested on the remaining studies, for every
study. A similar approach was then applied at the OTU level with the
exception that a 10-fold CV over 100 different models, based on random
80/20 splitting of the data, was used to generate a range of expected
AUCs. For these OTU based models, the CRC-associated genera included all
OTUs that had a taxonomic classification to \emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas}.
Finally, the power of each study was assessed for an effect size ranging
from 1\% to 30\% and an estimated sample size, for these effect sizes,
was generated based on 80\% power. For comparisons in which normal
versus adenoma were made the carcinoma samples were excluded from each
respective study. Similarly, for comparisons in which normal versus
carcinoma were made the adenoma samples were excluded from each
respective study. The data was split between feces and tissue samples.
Within the tissue groups the data was further divided between samples
from the same individual (matched) and those from different individuals
(unmatched). Where applicable for each study, predictions for adenoma
and carcinoma were then tested for feces, matched tissue, and unmatched
tissue.

\textbf{\emph{Obtaining CRC-associated Genera:}} For the CRC-associated
genera analysis of the RR, the total average counts, for 100 different
subsamplings, were collected for each respective OTU that had a genus
level taxonomic classification to \emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, and \emph{Porphyromonas}.
The OTU based Random Forest Models using CRC-associated genera utilized
a similar approach except that the OTUs were not aggregated together by
genus but kept as separate OTUs. OTU Random Forest models using the full
community included all OTUs while those using CRC-associated genera
included only those OTUs that had a genus level taxonomic classification
to \emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus},
and \emph{Porphyromonas}.
